---
title: "Convergence check for phylogenetic reconstructions"
output: html_notebook
---

```{r}
library(convenience)
```

# Full alignment
## OBD dating
### Spinose

```{r}
output_S <- checkConvergence(list_files = list.files("~/bioclust/Forams/4-Phylogenetic_reconstruction/output/", pattern = "forams_Spinose.*.trees", full.names = T), control = makeControl(burnin = 0.1))
output_S
```

```{r fig.width=12, fig.asp=0.4}
plotEssSplits(output_S, breaks=10)
#plotEssSplits(output_S, per_run = T, breaks = 10)
```

### Non-Spinose

```{r}
output_NS <- checkConvergence(list_files = list.files("~/bioclust/Forams/4-Phylogenetic_reconstruction/output", pattern = "forams_NonSpinose.*.trees", full.names = T), control = makeControl(burnin = 0.1))
output_NS
```

```{r fig.width=12, fig.asp=0.4}
plotEssSplits(output_NS, breaks = 20, )
#plotEssSplits(output_NS, per_run = T, breaks = 20)
```

### Microperforate

```{r}
output_M <- checkConvergence(list_files = list.files("~/bioclust/Forams/4-Phylogenetic_reconstruction/output/", pattern = "forams_Microperforate.*.trees", full.names = T), control = makeControl(burnin = 0.05))
output_M
```

```{r fig.width=12, fig.asp=0.4}
plotEssSplits(output_M, breaks=20)
#plotEssSplits(output_M, per_run = T, breaks=20)
```

# High-confidence alignment
## OBD dating
### Spinose

```{r}
output_95_S <- checkConvergence(list_files = list.files("~/bioclust/Forams/4-Phylogenetic_reconstruction/output_pf95_pt95/", pattern = "forams_Spinose.*.trees", full.names = T), control = makeControl(burnin = 0.1))
output_95_S
```

```{r fig.width=12, fig.asp=0.4}
plotEssSplits(output_95_S, breaks=10)
#plotEssSplits(output_95_S, per_run = T, breaks = 10)
```

### Non-Spinose

```{r}
output_95_NS <- checkConvergence(list_files = list.files("~/bioclust/Forams/4-Phylogenetic_reconstruction/output_pf95_pt95/", pattern = "forams_NonSpinose.*.trees", full.names = T), control = makeControl(burnin = 0.1))
output_95_NS
```

```{r fig.width=12, fig.asp=0.4}
plotEssSplits(output_95_NS)
#plotEssSplits(output_95_NS, per_run = T, breaks = 20)
```

### Microperforate

```{r}
output_95_M <- checkConvergence(list_files = list.files("~/bioclust/Forams/4-Phylogenetic_reconstruction/output_pf95_pt95/", pattern = "forams_Microperforate.*.trees", full.names = T), control = makeControl(burnin = 0.05))
output_95_M
```

```{r fig.width=12, fig.asp=0.4}
plotEssSplits(output_95_M, breaks=20)
#plotEssSplits(output_95_M, per_run = T, breaks=20)
```

## OBD + Node dating
### Spinose

```{r}
output_95_nodeDating_S <- checkConvergence(list_files = list.files("~/bioclust/Forams/4-Phylogenetic_reconstruction/output_pf95_pt95_nodeDating/", pattern = "forams_Spinose.*.trees", full.names = T), control = makeControl(burnin = 0.1))
output_95_nodeDating_S
```

```{r fig.width=12, fig.asp=0.4}
plotEssSplits(output_95_nodeDating_S, breaks=30)
plotEssSplits(output_95_nodeDating_S, per_run = T, breaks = 10)
```

### Non-Spinose

```{r}
output_95_nodeDating_NS <- checkConvergence(list_files = list.files("~/bioclust/Forams/4-Phylogenetic_reconstruction/output_pf95_pt95_nodeDating/", pattern = "forams_NonSpinose.*.trees", full.names = T), control = makeControl(burnin = 0.1))
output_95_nodeDating_NS
```

```{r fig.width=12, fig.asp=0.4, warning=FALSE}
plotEssSplits(output_95_nodeDating_NS)
plotEssSplits(output_95_nodeDating_NS, per_run = T, breaks = 20)
```

### Microperforate

```{r}
output_95_nodeDating_M <- checkConvergence(list_files = list.files("~/bioclust/Forams/4-Phylogenetic_reconstruction/output_pf95_pt95_nodeDating/", pattern = "forams_Microperforate.*.trees", full.names = T), control = makeControl(burnin = 0.05))
output_95_nodeDating_M
```

```{r fig.width=12, fig.asp=0.4}
plotEssSplits(output_95_nodeDating_M, breaks=20)
#plotEssSplits(output_95_nodeDating_M, per_run = T, breaks=20)
```

# Medium-confidence alignment
## OBD dating
### Spinose

```{r}
output_5_S <- checkConvergence(list_files = list.files("~/bioclust/Forams/4-Phylogenetic_reconstruction/output_pf5_pt5/", pattern = "forams_Spinose.*.trees", full.names = T), control = makeControl(burnin = 0.1))
output_5_S
```

```{r fig.width=12, fig.asp=0.4}
plotEssSplits(output_5_S, breaks=10)
```

### Non-Spinose

```{r}
output_5_NS <- checkConvergence(list_files = list.files("~/bioclust/Forams/4-Phylogenetic_reconstruction/output_pf5_pt5/", pattern = "forams_NonSpinose.*.trees", full.names = T), control = makeControl(burnin = 0.1))
output_5_NS
```

```{r fig.width=12, fig.asp=0.4, warning=FALSE}
plotEssSplits(output_5_NS)
plotEssSplits(output_5_NS, per_run = T, breaks = 20)
```

### Microperforate

```{r}
output_5_M <- checkConvergence(list_files = list.files("~/bioclust/Forams/4-Phylogenetic_reconstruction/output_pf5_pt5/", pattern = "forams_Microperforate.*.trees", full.names = T), control = makeControl(burnin = 0.05))
output_5_M
```

```{r fig.width=12, fig.asp=0.4}
plotEssSplits(output_5_M, breaks=50)
```

## OBD + Node dating
### Spinose

```{r}
output_5_nodeDating_S <- checkConvergence(list_files = list.files("~/bioclust/Forams/4-Phylogenetic_reconstruction/output_pf5_pt5_nodeDating/", pattern = "forams_Spinose.*.trees", full.names = T), control = makeControl(burnin = 0.1))
output_5_nodeDating_S
```

```{r fig.width=12, fig.asp=0.4}
plotEssSplits(output_5_nodeDating_S, breaks=30)
plotEssSplits(output_5_nodeDating_S, per_run = T, breaks = 20)
```

### Non-Spinose

```{r}
output_5_nodeDating_NS <- checkConvergence(list_files = list.files("~/bioclust/Forams/4-Phylogenetic_reconstruction/output_pf5_pt5_nodeDating/", pattern = "forams_NonSpinose.*.trees", full.names = T), control = makeControl(burnin = 0.1))
output_5_nodeDating_NS
```

```{r fig.width=12, fig.asp=0.4}
plotEssSplits(output_5_nodeDating_NS)
plotEssSplits(output_5_nodeDating_NS, per_run = T, breaks = 20)
```

### Microperforate

```{r}
output_5_nodeDating_M <- checkConvergence(list_files = list.files("~/bioclust/Forams/4-Phylogenetic_reconstruction/output_pf5_pt5_nodeDating/", pattern = "forams_Microperforate.*.trees", full.names = T), control = makeControl(burnin = 0.05))
output_5_nodeDating_M
```

```{r fig.width=12, fig.asp=0.4}
plotEssSplits(output_5_nodeDating_M, breaks=50)
```


# Summary: ESS quantiles across variants and clades

The default `checkConvergence` printout shows only the lowest split ESS per run. The summary below reports the 10th-percentile, median and 90th-percentile ESS across splits per (variant × clade × run), which is more informative for assessing convergence.

```{r}
ess_summary <- function(out) {
  ess <- out$tree_parameters$ess
  do.call(rbind, lapply(seq_len(ncol(ess)), function(r) {
    v <- as.numeric(ess[, r]); v <- v[is.finite(v)]
    data.frame(run = r, n_splits = length(v),
               p10 = round(quantile(v, 0.1), 1),
               median = round(median(v), 1),
               p90 = round(quantile(v, 0.9), 1))
  }))
}

outs <- list(
  `output / Spinose`                      = output_S,
  `output / NonSpinose`                   = output_NS,
  `output / Microperforate`               = output_M,
  `pf5_pt5 / Spinose`                     = output_5_S,
  `pf5_pt5 / NonSpinose`                  = output_5_NS,
  `pf5_pt5 / Microperforate`              = output_5_M,
  `pf95_pt95 / Spinose`                   = output_95_S,
  `pf95_pt95 / NonSpinose`                = output_95_NS,
  `pf95_pt95 / Microperforate`            = output_95_M,
  `pf95_pt95_nodeDating / Spinose`        = output_95_nodeDating_S,
  `pf95_pt95_nodeDating / NonSpinose`     = output_95_nodeDating_NS,
  `pf95_pt95_nodeDating / Microperforate` = output_95_nodeDating_M,
  `pf5_pt5_nodeDating / Spinose`          = output_5_nodeDating_S,
  `pf5_pt5_nodeDating / NonSpinose`       = output_5_nodeDating_NS,
  `pf5_pt5_nodeDating / Microperforate`   = output_5_nodeDating_M
)
ess_tab <- do.call(rbind, lapply(names(outs), function(k) cbind(label = k, ess_summary(outs[[k]]))))
ess_tab
```

The OBDD pipeline consumes `output_pf95_pt95_nodeDating/`, where convergence is good for Microperforate (median ESS ~18 000) and NonSpinose (median ~2 900) and marginal for Spinose (median ~820, p10 ~340).
