---
title: "Subsampling the Triton datase - Geographical rarefaction"
output: html_notebook
---

```{r}
library(divvy)
library(ggplot2)
library(gridExtra)
#library(units)
library(sf)
library(terra)
#library(vegan)
#library(iNEXT)
library(rnaturalearth)
#library(rnaturalearthdata)
```

```{r}
triton.ss.STT <- read.csv("../../3-Data_processed/Triton_occurrences/TritonDB_subsampled_STT_91253foss.csv", row.names = 1)
head(triton.ss.STT)
```

# Inspect the dataset
## Geographical data

```{r}
# Define coordinate columns
xyCartes <- c('longitude','latitude')  # Longitude and Latitude columns
xyCell   <- c('cellX','cellY')       # To store grid cell coordinates
```

```{r}
# Initialize Equal Earth projection
prj <- 'EPSG:8857'  # Equal Earth projection

# Create an empty raster with Equal Earth projection and appropriate resolution
# Adjust 'res' parameter as needed (e.g., 200000 meters ~ 2 degrees)
rWorld <- rast()
rPrj <- project(rWorld, prj, res = 50000)
values(rPrj) <- 1:ncell(rPrj)
```

```{r}
# Convert occurrence points to spatial features
occurrences <- vect(triton.ss.STT, geom = xyCartes, crs = 'epsg:4326')
occurrences_prj <- project(occurrences, prj)

# Assign each occurrence to a grid cell
triton.ss.STT$cell <- cells(rPrj, occurrences_prj)[, 'cell']
triton.ss.STT[, xyCell] <- xyFromCell(rPrj, triton.ss.STT$cell)

# Remove occurrences with NA cells (if any)
triton.ss.STT <- triton.ss.STT[!is.na(triton.ss.STT$cell), ]
```

## Plotting

```{r}
# Ensure uniqueness of localities based on grid cells
# This step is crucial to avoid inflating site numbers due to minor coordinate discrepancies
triton.ss.STT_unique <- uniqify(triton.ss.STT, xyCell)

# Convert unique occurrences to sf object for visualization
triton.ss.STT_unique_sf <- st_as_sf(triton.ss.STT_unique, coords = xyCell, crs = prj)
```

```{r}
# Load world map for plotting
world <- ne_countries(scale = "medium", returnclass = "sf")

# Visualize the current spatial distribution of subsampled data
world_plot <- ggplot(data = world) +
  theme_bw() +
  geom_sf() +
  geom_sf(data = triton.ss.STT_unique_sf, shape = 17, color = 'blue') +
  labs(title = "Spatial Distribution of Subsampled Triton Fossils")

print(world_plot)
```

# Perform Circular Subsampling Using 'cookies'
## Circular subsampling
Parameters:
- iter: Number of iterations (e.g., 500)
- nSite: Number of sites per subsample (e.g., 20)
- r: Radius in kilometers (e.g., 1500 km)

```{r}
set.seed(123)  # For reproducibility
circ_subsamples <- cookies(
  dat = triton.ss.STT_unique,  
  xy = xyCell,
  iter = 200, 
  nSite = 2,          # Adjust based on your dataset size and desired subsample size
  r = 1500,           # Radius in kilometers
  weight = TRUE,      # Probabilistically aggregate subsampling sites
  crs = prj, 
  output = 'full'     # Retrieve full occurrence data for each subsample
)

head(circ_subsamples[[1]])
```

```{r}
# Visualize the first subsample on the world map
first_subsample <- circ_subsamples[[sample(1:length(circ_subsamples), 1)]]
subsample_sf <- st_as_sf(first_subsample, coords = xyCell, crs = prj)

world_plot +
  geom_sf(data = subsample_sf, shape = 17, color = 'red') +
  labs(title = "First Circular Subsample of Triton Fossils")

# Visualize the circular boundary around the seed location
seed_location <- subsample_sf[1, ]
buffer_radius <- 1500 * 1000  # Convert km to meters
buffer <- st_buffer(seed_location, dist = buffer_radius)

# Plot with buffer
world_plot +
  geom_sf(data = subsample_sf, shape = 17, color = 'red') +
  geom_sf(data = buffer, fill = NA, color = 'red', size = 1) +
  labs(title = "First Circular Subsample with Boundary")
```

# Combine All Sampled Sites

```{r}
# Combine all subsamples into a single data frame
all_circ_subsamples <- do.call(rbind, circ_subsamples)

# Ensure that the combined data frame has unique sites
all_circ_unique <- uniqify(all_circ_subsamples, xyCell)

# Extract unique sampled sites across all iterations
sampled_sites <- all_circ_unique$cell
```

## Compare Geographical Distribution Before and After Subsampling

```{r}
# Identify sampled sites by their grid cells
triton.ss.STT_unique_sf$sampled <- triton.ss.STT_unique_sf$cell %in% sampled_sites

# Create a new column for color based on sampling
triton.ss.STT_unique_sf$color <- ifelse(triton.ss.STT_unique_sf$sampled, "Sampled", "Not Sampled")

# Plot the map with sampled sites in red and others in blue
ggplot(data = world) +
  theme_bw() +
  geom_sf() +
  geom_sf(data = triton.ss.STT_unique_sf, aes(color = color), shape = 17, alpha = 0.5) +
  scale_color_manual(values = c("Sampled" = "red", "Not Sampled" = "blue")) +
  labs(title = "Geographical Subsampling of Triton Fossils",
       color = "Site Status") +
  theme(legend.position = "bottom")
```

# Compare Latitude and Longitude Densities
In this section, we will compare the density distributions of latitude and longitude between the original and subsampled Triton fossil datasets. This comparison helps assess how well the subsampling process has standardized the spatial distribution of fossil occurrences.

```{r}
# Calculate Density Estimates
triton.ss.STT.Geo <- triton.ss.STT[triton.ss.STT$cell %in% sampled_sites,]

density_original_lat <- density(triton.ss.STT$latitude, from = -90, to = 90, na.rm = TRUE)
density_sampled_lat <- density(triton.ss.STT.Geo$latitude, from = -90, to = 90, na.rm = TRUE)

density_original_long <- density(triton.ss.STT$longitude, from = -180, to = 180, na.rm = TRUE)
density_sampled_long <- density(triton.ss.STT.Geo$longitude, from = -180, to = 180, na.rm = TRUE)
```

```{r fig.width=12, fig.asp=0.3}
# Latitude Density Comparison
p1 <- ggplot() +
  geom_line(aes(x = density_original_lat$x, y = density_original_lat$y, color = "Original"), size = 1) +
  geom_line(aes(x = density_sampled_lat$x, y = density_sampled_lat$y, color = "Sampled"), size = 1, linetype = "dashed") +
  scale_color_manual(values = c("Original" = "blue", "Sampled" = "red")) +
  labs(title = "Latitude Density Comparison",
       x = "Latitude",
       y = "Density",
       color = "") +
  theme_minimal()

# Longitude Density Comparison
p2 <- ggplot() +
  geom_line(aes(x = density_original_long$x, y = density_original_long$y, color = "Original"), size = 1) +
  geom_line(aes(x = density_sampled_long$x, y = density_sampled_long$y, color = "Sampled"), size = 1, linetype = "dashed") +
  scale_color_manual(values = c("Original" = "blue", "Sampled" = "red")) +
  labs(title = "Longitude Density Comparison",
       x = "Longitude",
       y = "Density",
       color = "") +
  theme_minimal()

# Arrange both plots side by side
grid.arrange(p1, p2, ncol = 2)
```

```{r}
hist(sapply(na.omit(unique(triton.ss.STT$holeID)), function(holeID)triton.ss.STT[na.omit(triton.ss.STT$holeID == holeID),"latitude"][1]), breaks=40)
hist(sapply(na.omit(unique(triton.ss.STT$holeID)), function(holeID)triton.ss.STT[na.omit(triton.ss.STT$holeID == holeID),"longitude"][1]), breaks=20)
hist(sapply(na.omit(unique(triton.ss.STT.Geo$holeID)), function(holeID)triton.ss.STT.Geo[na.omit(triton.ss.STT.Geo$holeID == holeID),"latitude"][1]), breaks=40)
hist(sapply(na.omit(unique(triton.ss.STT.Geo$holeID)), function(holeID)triton.ss.STT.Geo[na.omit(triton.ss.STT.Geo$holeID == holeID),"longitude"][1]), breaks=20)
```


# Extract all subsampled datasets
# Each element in 'circ_subsamples' is a subsampled dataset

```{r}
# Combine all subsamples into a list of data frames
all_circ_subsamples <- lapply(circ_subsamples, as.data.frame)
```

# Analyze Subsample Distributions (Optional)
# For example, compare the latitudinal and longitudinal distributions

```{r}
# Function to calculate density for a subsample
calculate_density <- function(subsample, type = c("lat", "long")){
  type <- match.arg(type)
  if(type == "lat"){
    return(density(subsample$pal.lat, from = -90, to = 90, na.rm = TRUE))
  } else {
    return(density(subsample$pal.long, from = -180, to = 180, na.rm = TRUE))
  }
}
```

```{r}
# Plot example density comparison
par(mfrow = c(1,2))
plot(calculate_density(triton.ss.STT_unique, "lat"), main = "Latitude Density", xlab = "Latitude", ylab = "Density")
for(i in 1:100){
  lines(calculate_density(all_circ_subsamples[[i]], "lat"), col = alpha("red", 0.3))
}

plot(calculate_density(triton.ss.STT_unique, "long"), main = "Longitude Density", xlab = "Longitude", ylab = "Density")
for(i in 1:100){
  lines(calculate_density(all_circ_subsamples[[i]], "long"), col = alpha("blue", 0.3))
}
```
