---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r include=FALSE}
library("ggplot2")
library("divDyn")
library("rEDM")
#library("sjPlot")
```

# Load processed data

```{r}
triton <- read.csv("../../3-Data_processed/Triton_occurrences/TritonDB_trimmed_runif.csv", row.names=1)
triton.ss <- read.csv("../../3-Data_processed/Triton_occurrences/TritonDB_subsampled_STT_91253foss.csv", row.names=1)
```

```{r}
time_points <- read.csv("../../3-Data_processed/Triton_occurrences/time_points.csv")[,1]
#OTT <- read.csv("../../3-Data_processed/Triton_occurrences/OTT.csv")[,1]
expertLTT <- read.csv("../../3-Data_processed/Triton_occurrences/expertLTT.csv")[,1]
#empiricalLTT <- read.csv("../../3-Data_processed/Triton_occurrences/empiricalLTT.csv")[,1]
#OTT.ss <- read.csv("../../3-Data_processed/Triton_occurrences/OTT_ss.csv")[,1]
#expertLTT.ss <- read.csv("../../3-Data_processed/Triton_occurrences/expertLTT_ss.csv")[,1]
#empiricalLTT.ss <- read.csv("../../3-Data_processed/Triton_occurrences/empiricalLTT_ss.csv")[,1]
```

# DivDyn plots

Classical methods to estimate diversification rates from the fossil record.

```{r}
data(tens)
data(stages)
data(keys)
# names of the bins
colnames(tens)[colnames(tens)=="X10"] <- "name"
colnames(stages)[colnames(stages)=="stage"] <- "name"
tens
stages
```

Select columns from the Triton database. Diversification analysis will be run using `age_runif`, sampled within the fossil age uncertainty range.

```{r}
triton_divDyn <- triton[c("species", "age", "age_runif", "holeID")]

# 1My bins
triton_divDyn$MyrBin <- round(triton_divDyn$age_runif)
sampMyrBin <- binstat(triton_divDyn, tax="species", bin="MyrBin", coll="holeID", duplicates=FALSE)

# Stage bins (homogenize names)
triton_divDyn$stage <- sapply(triton_divDyn$age_runif, function(ti)stages[stages$bottom>ti & ti>=stages$top,]$name)
triton_divDyn$stage <- gsub("Lower Miocene", "Early Miocene", triton_divDyn$stage)
triton_divDyn$stage <- gsub("Upper Miocene", "Late Miocene", triton_divDyn$stage)
triton_divDyn$stage <- gsub("Selandian-Thanetian", "Selandian", triton_divDyn$stage)
triton_divDyn$stg <- as.numeric(categorize(triton_divDyn$stage, keys$stgInt))
sampStg <- binstat(triton_divDyn, tax="species", bin="stg", coll="holeID", noNAStart=FALSE, duplicates=FALSE)
```

```{r fig.asp=0.4, fig.width=12}
par(mfrow=c(1,2))
# Plot every stage
tsplot(stages, shading="short", boxes=c("short", "sys"), boxes.col=c("col", "systemCol"), xlim=c(66,0), ylim=c(0,15000), labels.args=list(cex=0.7), ylab="Occurrences (black) and collections (blue)"); title("Foraminiferan fossils at each geological stage")
# occurrences
lines(stages$mid, sampStg$occs, lwd=2)
# collections
lines(stages$mid, sampStg$colls, lwd=2, col="blue")

# Plot every 1Myr
tsplot(stages, shading="short", boxes=c("short", "sys"), boxes.col=c("col", "systemCol"), xlim=c(66,0), ylim=c(0,15000), labels.args=list(cex=0.7), ylab="Occurrences (black) and collections (blue)"); title("Foraminiferan fossils at each 1Myr time bin")
# occurrences
lines(sampMyrBin$MyrBin, sampMyrBin$occs, lwd=2)
# collections
lines(sampMyrBin$MyrBin, sampMyrBin$colls, lwd=2, col="blue")
```

```{r fig.asp=0.5, fig.width=12}
tsplot(stages, shading="short", boxes=c("short", "sys"), boxes.col=c("col", "systemCol"), xlim=c(66,0), labels.args=list(cex=0.7))
ranges(triton_divDyn, tax="species", bin="age_runif")
title("Stratigraphic ranges for foraminiferan species in the Triton database")
```

Run divDyn.

```{r}
ddStages <- divDyn(triton_divDyn, bin="stg", tax="species")
ddMyrBins <- divDyn(triton_divDyn, bin="MyrBin", tax="species")
```

```{r fig.asp=0.4, fig.width=12}
par(mfrow=c(1,2))

tsplot(stages, shading="short", boxes=c("short", "sys"), boxes.col=c("col", "systemCol"), xlim=c(66,0), ylim=c(0,300), labels.args=list(cex=0.7))
title("Sampled-in-bin foraminiferan diversity (stages)")
lines(stages$mid, ddStages$divCSIB, col="darkgreen", lwd=2)
lines(stages$mid, ddStages$divSIB, lwd=1)
legend("top", inset=c(0.01,0.01), legend=c("raw SIB", "corrected SIB"),
col=c("black", "darkgreen"), lwd=c(1,2), bg="white", cex=1)

tsplot(stages, shading="short", boxes=c("short", "sys"), boxes.col=c("col", "systemCol"), xlim=c(66,0), ylim=c(0,300), labels.args=list(cex=0.7))
title("Sampled-in-bin foraminiferan diversity (1Myr bins)")
lines(sampMyrBin$MyrBin, ddMyrBins$divCSIB, col="darkgreen", lwd=2)
lines(sampMyrBin$MyrBin, ddMyrBins$divSIB, lwd=1)
legend("top", inset=c(0.01,0.01), legend=c("raw SIB", "corrected SIB"), col=c("black", "darkgreen"), lwd=c(1,2), bg="white", cex=1)
```

100 iterations of Shareholder Quorum Subsampling (SQS).

```{r message=FALSE, warning=FALSE}
sqsMyrBinsPlot <- subsample(triton_divDyn, bin="MyrBin", tax="species", coll="holeID", q=0.7, iter=500, ref="holeID", singleton="ref", type="sqs", duplicates=FALSE, excludeDominant=TRUE, largestColl =TRUE, output="dist", na.rm=TRUE)
```

Diversity trends, using raw data, expert estimates (from the speciation and extinction dates in Triton) and mean of the SQS trials.

```{r fig.asp=0.6, fig.width=10}
tsplot(stages, shading="short", boxes=c("short", "sys"), boxes.col=c("col", "systemCol"), xlim=c(66,0), ylim=c(0,120), labels.args=list(cex=0.7))
title("Comparison between estimates of foram richness trends")
lines(sampMyrBin$MyrBin, ddMyrBins$divCSIB, col="darkgreen", lwd=2)
lines(time_points, expertLTT, col="purple", lwd=2)

# loop through all trial results
for(i in 1:ncol(sqsMyrBinsPlot$divCSIB)){
  lines(sampMyrBin$MyrBin, sqsMyrBinsPlot$divCSIB[,i], col="#00000005", lwd=1)
}
legend("topleft", bg="white", legend=c("raw data (CSIB)", "experts' LTT", "SQS trials"), col=c("darkgreen", "purple", "#00000088"), lwd=3, inset=c(0.01, 0.01, 0.01))

#the mean of the trial results
meanSQS <- rowMeans(sqsMyrBinsPlot$divCSIB, na.rm = T)
lines(sampMyrBin$MyrBin, meanSQS, col="black", lwd=2)
```

Several methods to compute diversification rates, by increasing reliability (note that some give negative rates).

```{r fig.asp=0.6, fig.width=8}
tsplot(stages, shading="short", boxes=c("short", "sys"), boxes.col=c("col", "systemCol"), xlim=c(66,0), ylim=c(0,0.3), labels.args=list(cex=0.7))
title("Per capita rates")
lines(sampMyrBin$MyrBin, ddMyrBins$extPC, col="red", lwd=2)
lines(sampMyrBin$MyrBin, ddMyrBins$oriPC, col="blue", lwd=2)
legend("topright", legend=c("extinctions", "originations"), col=c("red", "blue"), lwd=2, bg="white", inset=c(0.01, 0.01))

tsplot(stages, shading="short", boxes=c("short", "sys"), boxes.col=c("col", "systemCol"), xlim=c(66,0), ylim=c(0,0.3), labels.args=list(cex=0.7))
title("Corrected three-timer rates")
lines(sampMyrBin$MyrBin, ddMyrBins$extC3t, col="red", lwd=2)
lines(sampMyrBin$MyrBin, ddMyrBins$oriC3t, col="blue", lwd=2)
legend("topright", legend=c("extinctions", "originations"), col=c("red", "blue"), lwd=2, bg="white", inset=c(0.01, 0.01))

tsplot(stages, shading="short", boxes=c("short", "sys"), boxes.col=c("col", "systemCol"), xlim=c(66,0), ylim=c(0, 0.3), labels.args=list(cex=0.7))
title("Gap-filler rates")
lines(sampMyrBin$MyrBin, ddMyrBins$extGF, col="red", lwd=2)
lines(sampMyrBin$MyrBin, ddMyrBins$oriGF, col="blue", lwd=2)
legend("topright", legend=c("extinctions", "originations"), col=c("red", "blue"), lwd=2, bg="white", inset=c(0.01, 0.01))

tsplot(stages, shading="short", boxes=c("short", "sys"), boxes.col=c("col", "systemCol"), xlim=c(66,0), ylim=c(0, 0.3), labels.args=list(cex=0.7))
title("Second-for-third substitution rates")
lines(sampMyrBin$MyrBin, ddMyrBins$ext2f3, col="red", lwd=2)
lines(sampMyrBin$MyrBin, ddMyrBins$ori2f3, col="blue", lwd=2)
legend("topright", legend=c("extinctions", "originations"), col=c("red", "blue"), lwd=2, bg="white", inset=c(0.01, 0.01))
```

First approach to include some uncertainty: envelope plot from all the methods.

```{r fig.asp=0.4, fig.width=12}
par(mfrow=c(1,2))

# extinctions
# 1st panel - 1myr bins
par(mar=c(2, 5.1, 4.1, 1)) # margin adjustment
# plot
tsplot(stages, shading="short", boxes=c("short", "sys"), boxes.col=c("col", "systemCol"), xlim=c(66,0), ylim=c(0, 0.3), labels.args=list(cex=0.7))
title("Extinction rates")
extMat <- as.matrix(ddMyrBins[grep("ext", names(ddMyrBins), value=T)])
shades(sampMyrBin$MyrBin, extMat, col="red") # background
for(i in 2:ncol(extMat)) lines(sampMyrBin$MyrBin, extMat[,i] , lwd=0.5) # extinctions - BIN

# extinctions
# 1st panel - 1myr bins
par(mar=c(2, 5.1, 4.1, 1)) # margin adjustment
# plot
tsplot(stages, shading="short", boxes=c("short", "sys"), boxes.col=c("col", "systemCol"), xlim=c(66,0), ylim=c(0, 0.3), labels.args=list(cex=0.7))
title("Origination rates")
oriMat <- as.matrix(ddMyrBins[grep("ori", names(ddMyrBins), value=T)])
shades(sampMyrBin$MyrBin, oriMat, col="blue") # background
for(i in 2:ncol(oriMat)) lines(sampMyrBin$MyrBin, oriMat[,i] , lwd=0.5) # originations - BIN
```

Second approach to include some uncertainty: envelope plot from all the SQS trials

```{r fig.asp=0.5, fig.width=12}
#par(mfrow=c(3,1))
layout(mat = matrix(c(1, 2, 3, 3), nrow = 2, byrow = F))

# extinction rates
par(mar=c(2, 5.1, 4.1, 1)) # margin adjustment
# plot
tsplot(stages, shading="short", boxes=c("short", "sys"), boxes.col=c("col", "systemCol"), xlim=c(66,0), ylim=c(0, 0.5), labels.args=list(cex=0.7))
title("Extinction rates")
shades(sampMyrBin$MyrBin, sqsMyrBinsPlot$ext2f3, col="red", res=50) # background
#for(i in 2:ncol(extMat)) lines(sampMyrBin$MyrBin, extMat[,i] , lwd=0.5) # extinctions - BIN

# origination rates
# plot
tsplot(stages, shading="short", boxes=c("short", "sys"), boxes.col=c("col", "systemCol"), xlim=c(66,0), ylim=c(0, 0.5), labels.args=list(cex=0.7))
title("Origination rates")
shades(sampMyrBin$MyrBin, sqsMyrBinsPlot$ori2f3, col="blue", res=50) # background
#for(i in 2:ncol(oriMat)) lines(sampMyrBin$MyrBin, oriMat[,i] , lwd=0.5) # originations - BIN

# diversification rates
# plot
tsplot(stages, shading="short", boxes=c("short", "sys"), boxes.col=c("col", "systemCol"), xlim=c(66,0), ylim=c(-0.3, 0.3), labels.args=list(cex=0.7))
title("Net-diversification rates")
shades(sampMyrBin$MyrBin, sqsMyrBinsPlot$ori2f3-sqsMyrBinsPlot$ext2f3, col="darkgreen", res=50) # background
#for(i in 2:ncol(oriMat)) lines(sampMyrBin$MyrBin, oriMat[,i] , lwd=0.5) # originations - BIN
```

Compute the mean speciation and extinction rates with the 2f3 method from the SQS trials.

```{r fig.asp=0.3, fig.width=7}
par(mfrow=c(1,2))
mean(sqsMyrBinsPlot$ori2f3, na.rm=T)
mean(sqsMyrBinsPlot$ext2f3, na.rm=T)
```

## Diversity dependance
### Plots and correlations

Relationship between speciation and extinction rates.

```{r fig.asp=0.4, fig.width=12}
par(mfrow=c(1,2))
plot(sqsMyrBinsPlot$ori2f3, sqsMyrBinsPlot$ext2f3, pch=3, cex=0.5, col=alpha("black",0.1), main="Relation between extinction and origination rates (2f3)", xlab="Origination rates (2f3)", ylab="Extinction rates (2f3)")
ori2f3_mean = rowMeans(sqsMyrBinsPlot$ori2f3)
ext2f3_mean = rowMeans(sqsMyrBinsPlot$ext2f3)
plot(ori2f3_mean, ext2f3_mean, pch=16, cex=0.5, col="red", xlim=range(sqsMyrBinsPlot$ori2f3, na.rm=T), ylim=range(sqsMyrBinsPlot$ext2f3, na.rm=T), main="Relation between mean extinction and origination rates (2f3)", xlab="Origination rates (2f3)", ylab="Extinction rates (2f3)")
ori2f3_sd = sqrt(rowMeans((sqsMyrBinsPlot$ori2f3-ori2f3_mean)**2))
ext2f3_sd = sqrt(rowMeans((sqsMyrBinsPlot$ext2f3-ext2f3_mean)**2))
arrows(x0=ori2f3_mean, x1=ori2f3_mean, y0=ext2f3_mean-ext2f3_sd, y1=ext2f3_mean+ext2f3_sd, code=3, angle=90, length=0.01, col=alpha("red", 0.2), lwd=0.5)
arrows(x0=ori2f3_mean-ori2f3_sd, x1=ori2f3_mean+ori2f3_sd, y0=ext2f3_mean, y1=ext2f3_mean, code=3, angle=90, length=0.01, col=alpha("red", 0.2), lwd=0.5)
```

```{r fig.asp=0.8, fig.width=10}
par(mfrow=c(2,2))
plot(sqsMyrBinsPlot$divCSIB, sqsMyrBinsPlot$ori2f3, pch=3, cex=0.5, col=alpha("black",0.1), main="Diversity-dependance on origination rates", xlab="Subsampled species richness (SQS)", ylab="Extinction rates (2f3)")
plot(sqsMyrBinsPlot$divCSIB, sqsMyrBinsPlot$ext2f3, pch=3, cex=0.5, col=alpha("black",0.1), main="Diversity-dependance on extinction rates", xlab="Subsampled species richness (SQS)", ylab="Extinction rates (2f3)")

divSQS_sd = sqrt(rowMeans((sqsMyrBinsPlot$divCSIB-meanSQS)**2))

plot(meanSQS, ori2f3_mean, pch=3, cex=0.5, col="red", main="Diversity-dependance on mean origination rates", xlab="Subsampled species richness (SQS)", ylab="Origination rates (2f3)", xlim=range(sqsMyrBinsPlot$divCSIB, na.rm=T), ylim=range(sqsMyrBinsPlot$ori2f3, na.rm=T))
arrows(x0=meanSQS, x1=meanSQS, y0=ori2f3_mean-ori2f3_sd, y1=ori2f3_mean+ori2f3_sd, code=3, angle=90, length=0.01, col=alpha("red", 0.2), lwd=0.5)
arrows(x0=meanSQS-divSQS_sd, x1=meanSQS+divSQS_sd, y0=ori2f3_mean, y1=ori2f3_mean, code=3, angle=90, length=0.01, col=alpha("red", 0.2), lwd=0.5)

plot(meanSQS, ext2f3_mean, pch=3, cex=0.5, col="red", main="Diversity-dependance on mean extinction rates", xlab="Subsampled species richness (SQS)", ylab="Extinction rates (2f3)", xlim=range(sqsMyrBinsPlot$divCSIB, na.rm=T), ylim=range(sqsMyrBinsPlot$ext2f3, na.rm=T))
arrows(x0=meanSQS, x1=meanSQS, y0=ext2f3_mean-ext2f3_sd, y1=ext2f3_mean+ext2f3_sd, code=3, angle=90, length=0.01, col=alpha("red", 0.2), lwd=0.5)
arrows(x0=meanSQS-divSQS_sd, x1=meanSQS+divSQS_sd, y0=ext2f3_mean, y1=ext2f3_mean, code=3, angle=90, length=0.01, col=alpha("red", 0.2), lwd=0.5)
```
Basically 0 correlation.

```{r}
summary(lm(ori2f3_mean ~ meanSQS))
summary(lm(ext2f3_mean ~ meanSQS))
```

### Lagged auto-correlations and cross-correlations

```{r fig.asp=0.6, fig.width=10}
par(mfrow=c(2,3), mar=c(5, 4, 4, 2))
acf(ori2f3_mean, main="Auto-Correlation - Origination rates", na.action = na.pass)
acf(ext2f3_mean, main="Auto-Correlation - Extinction rates", na.action = na.pass)
acf(meanSQS, main="Auto-Correlation - SQS richness", na.action = na.pass)
ccf(ext2f3_mean, ori2f3_mean, main="Cross-Correlation - Ori vs Ext rates", na.action = na.pass)
ccf(ori2f3_mean, meanSQS, main="Cross-Correlation - Ori vs SQS richness", na.action = na.pass)
ccf(ext2f3_mean, meanSQS, main="Cross-Correlation - Ext vs SQS richness", na.action = na.pass)
```

Observations:
1. Diversification rates and SQS richness are all somewhat autocorrelated.
2. Origination rates are positively cross-correlated with contemporaneous and previous (1My older) extinction rates.
3. Extinction rates are negatively cross-correlated with previous (1Myr and 2Myr older) diversity.

### SDEs : layeranalyzer

Reitan, T, Liow, LH. layeranalyzer: Inferring correlative and causal connections from time series data in r. Methods Ecol Evol. 2019 https://doi.org/10.1111/2041-210X.13299

```{r}
#devtools::install_github(repo="trondreitan/layeranalyzer",dependencies=TRUE,build_vignettes=TRUE)
#library("layeranalyzer")

#models=traverse.standalone.layered(hare.norm, max.layers=2, 
#  talkative=TRUE, allow.one.feedback.loop=TRUE, 
#  just.stationary=FALSE, no.rw=FALSE,    
#  time.integrals.possible=FALSE, 
#  allow.deterministic.layers=TRUE,
#  use.stationary.stdev = FALSE, mcmc=TRUE,
#  num.MCMC=1000,spacing=10,burnin=2000, num.temp = 4, prior=p.hare)
#compare.layered(models)
```

### Convergent cross mapping

Empirical Dynamic Modeling (2022-03-31)
George Sugihara, Joseph Park, Ethan Deyle, Erik Saberski, Cameron Smith, Hao Ye
https://cran.gedik.edu.tr/web/packages/rEDM/vignettes/rEDM-tutorial.pdf

First, select the embedding dimension using a nearest neighbor forecasting method:
`lib` = “training” portion of the data, on which nearest neighbors can be identified
`pred` = “test” portion of the data, on which 1-step ahead forecasts will be made

```{r fig.asp=0.3, fig.width=12}
rates_2f3 <- data.frame(test=1, ori=ori2f3_mean, ori_sd=ori2f3_sd, ext=ext2f3_mean, ext_sd=ext2f3_sd)
par(mfrow=c(1,3))

get_rho_E_matrix <- function(t_series_1, t_series_2){
  return (sapply(1:ncol(t_series_1), function(i){
    rates_2f3_i <- data.frame(test=1, col=t_series_1[,i], target=t_series_2[,i])
    EmbedDimension(dataFrame=na.omit(rates_2f3_i), lib="1 30", pred="30 63", columns="col", target = "target", showPlot=F)$rho
  }))
}

rho_E_matrix <- get_rho_E_matrix(sqsMyrBinsPlot$ori2f3, sqsMyrBinsPlot$ori2f3)
matplot(rho_E_matrix, type="l", col=alpha("black", 0.02), lwd=2, lty=1, xlab="Embeding dimension E", ylab="Prediction Skill", main="Selecting E for origination")
lines(apply(rho_E_matrix, 1, median), col="black", lwd=2)
paste("Optimal E value for origination rates:", which.max(apply(rho_E_matrix, 1, median)))

rho_E_matrix <- get_rho_E_matrix(sqsMyrBinsPlot$ext2f3, sqsMyrBinsPlot$ext2f3)
matplot(rho_E_matrix, type="l", col=alpha("black", 0.02), lwd=2, lty=1, xlab="Embeding dimension E", ylab="Prediction Skill", main="Selecting E for extinction")
lines(apply(rho_E_matrix, 1, median), col="black", lwd=2)
paste("Optimal E value for extinction rates:", which.max(apply(rho_E_matrix, 1, median)))

rho_E_matrix <- get_rho_E_matrix(sqsMyrBinsPlot$ori2f3, sqsMyrBinsPlot$ext2f3)
matplot(rho_E_matrix, type="l", col=alpha("black", 0.02), lwd=2, lty=1, xlab="Embeding dimension E", ylab="Prediction Skill", main="Selecting E for origination vs extinction")
lines(apply(rho_E_matrix, 1, median), col="black", lwd=2)
paste("Optimal E value for origination vs extinction rates:", which.max(apply(rho_E_matrix, 1, median)))
```

The median results are very flat, there probably isn't much signal in this data.

```{r fig.asp=0.3, fig.width=12}
par(mfrow=c(1,3))
get_rho_Tp_matrix <- function(t_series_1, t_series_2, E){
  return (sapply(1:ncol(t_series_1), function(i){
    rates_2f3_i <- data.frame(test=1, col=t_series_1[,i], target=t_series_2[,i])
    PredictInterval(dataFrame=na.omit(rates_2f3_i), lib="1 30", pred="30 63", columns="col", target="target", E=E, showPlot=F)$rho
  }))
}

rho_Tp_matrix <- get_rho_Tp_matrix(sqsMyrBinsPlot$ori2f3, sqsMyrBinsPlot$ori2f3, E=7)
paste("Optimal Tp value for origination rates:", which.max(apply(rho_Tp_matrix, 1, median)))
matplot(rho_Tp_matrix, type="l", col=alpha("black", 0.02), lwd=2, lty=1, xlab="Forecast Interval Tp", ylab="Prediction Skill", main="Selecting Tp for origination")
lines(apply(rho_Tp_matrix, 1, median), col="black", lwd=2)

rho_Tp_matrix <- get_rho_Tp_matrix(sqsMyrBinsPlot$ext2f3, sqsMyrBinsPlot$ext2f3, E=3)
paste("Optimal Tp value for extinction rates:", which.max(apply(rho_Tp_matrix, 1, median)))
matplot(rho_Tp_matrix, type="l", col=alpha("black", 0.02), lwd=2, lty=1, xlab="Forecast Interval Tp", ylab="Prediction Skill", main="Selecting Tp for extinction")
lines(apply(rho_Tp_matrix, 1, median), col="black", lwd=2)

rho_Tp_matrix <- get_rho_Tp_matrix(sqsMyrBinsPlot$ori2f3, sqsMyrBinsPlot$ext2f3, E=6)
paste("Optimal Tp value for origination vs extinction rates:", which.max(apply(rho_Tp_matrix, 1, median)))
matplot(rho_Tp_matrix, type="l", col=alpha("black", 0.02), lwd=2, lty=1, xlab="Forecast Interval Tp", ylab="Prediction Skill", main="Selecting Tp for origination vs extinction")
lines(apply(rho_Tp_matrix, 1, median), col="black", lwd=2)
```

```{r fig.asp=0.3, fig.width=12}
par(mfrow=c(1,3))
get_rho_Tp_matrix_Nonlinear <- function(t_series_1, t_series_2, E){
  return (sapply(1:ncol(t_series_1), function(i){
    rates_2f3_i <- data.frame(test=1, col=t_series_1[,i], target=t_series_2[,i])
    PredictNonlinear(dataFrame=na.omit(rates_2f3_i), lib="1 30", pred="30 63", columns="col", target="target", E=E, showPlot=F)$rho
  }))
}

rho_Tp_matrix <- get_rho_Tp_matrix_Nonlinear(sqsMyrBinsPlot$ori2f3, sqsMyrBinsPlot$ori2f3, E=7)
paste("Optimal S−map Localisation for origination rates:", which.max(apply(rho_Tp_matrix, 1, median)))
matplot(rho_Tp_matrix, type="l", col=alpha("black", 0.02), lwd=2, lty=1, xlab="S−map Localisation", ylab="Prediction Skill", main="Selecting S−map Localisation for origination")
lines(apply(rho_Tp_matrix, 1, median), col="black", lwd=2)

rho_Tp_matrix <- get_rho_Tp_matrix_Nonlinear(sqsMyrBinsPlot$ext2f3, sqsMyrBinsPlot$ext2f3, E=3)
paste("Optimal S−map Localisation for extinction rates:", which.max(apply(rho_Tp_matrix, 1, median)))
matplot(rho_Tp_matrix, type="l", col=alpha("black", 0.02), lwd=2, lty=1, xlab="S−map Localisation", ylab="Prediction Skill", main="Selecting S−map Localisation for extinction")
lines(apply(rho_Tp_matrix, 1, median), col="black", lwd=2)

rho_Tp_matrix <- get_rho_Tp_matrix_Nonlinear(sqsMyrBinsPlot$ori2f3, sqsMyrBinsPlot$ext2f3, E=6)
paste("Optimal S−map Localisation for origination vs extinction rates:", which.max(apply(rho_Tp_matrix, 1, median)))
matplot(rho_Tp_matrix, type="l", col=alpha("black", 0.02), lwd=2, lty=1, xlab="S−map Localisation", ylab="Prediction Skill", main="Selecting S−map Localisation for origination vs extinction")
lines(apply(rho_Tp_matrix, 1, median), col="black", lwd=2)
```

Now perform the convergent cross-mapping with the selected parameters.

`sample` = number of subsamples generated at each library size

Results on a single rate trajectory (probably not reliable):

```{r warning=FALSE, fig.asp=0.3, fig.width=10, message=FALSE, warning=FALSE}
par(mfrow=c(1,3))
cmap <- CCM(dataFrame=na.omit(rates_2f3), E=7, Tp=7, columns="ori", target="ori", libSizes = "7 30 1", sample=100, showPlot = TRUE)
cmap <- CCM(dataFrame=na.omit(rates_2f3), E=3, Tp=9, columns="ext", target="ext", libSizes = "3 30 1", sample = 100, showPlot = TRUE)
cmap <- CCM(dataFrame=na.omit(rates_2f3), E=6, Tp=1, columns="ori", target="ext", libSizes = "6 30 1", sample = 100, showPlot = TRUE)
```

Observations:
  1. Negative values for ori:ori -> no prediction skill -> no causal effect
  2. Negative values for ori:ori -> no prediction skill -> no causal effect
  3. Negative values for ori:ext -> no prediction skill -> no causal effect
  4. Positive values for ext:ori -> prediction skills -> origination rates influence extinction rates

Results on all SQS samples:

```{r fig.asp=0.3, fig.width=5, message=FALSE, warning=FALSE}
get_CCM_matrix <- function(t_series_1, t_series_2, E, Tp){
  return (sapply(1:ncol(t_series_1), function(i){
    rates_2f3_i <- data.frame(test=1, col=t_series_1[,i], target=t_series_2[,i])
    CCM(dataFrame=na.omit(rates_2f3_i), columns="col", target="target", libSizes=paste(E,"30 1"), E=E, Tp=Tp, sample=100, showPlot=F)
  }))
}

CCM_matrix <- get_CCM_matrix(sqsMyrBinsPlot$ori2f3, sqsMyrBinsPlot$ori2f3, E=7, Tp=7)
CCM_matrix_LibSize <- sapply(1:ncol(CCM_matrix), function(i)CCM_matrix[,i]$LibSize)
CCM_matrix_ColTarget <- sapply(1:ncol(CCM_matrix), function(i)CCM_matrix[,i]$`col:target`)
CCM_matrix_TargetCol <- sapply(1:ncol(CCM_matrix), function(i)CCM_matrix[,i]$`target:col`)

CCM_matrix <- get_CCM_matrix(sqsMyrBinsPlot$ext2f3, sqsMyrBinsPlot$ext2f3, E=3, Tp=9)
CCM_matrix_LibSize <- sapply(1:ncol(CCM_matrix), function(i)CCM_matrix[,i]$LibSize)
CCM_matrix_ColTarget <- sapply(1:ncol(CCM_matrix), function(i)CCM_matrix[,i]$`col:target`)
CCM_matrix_TargetCol <- sapply(1:ncol(CCM_matrix), function(i)CCM_matrix[,i]$`target:col`)

CCM_matrix <- get_CCM_matrix(sqsMyrBinsPlot$ori2f3, sqsMyrBinsPlot$ext2f3, E=6, Tp=1)
CCM_matrix_LibSize <- sapply(1:ncol(CCM_matrix), function(i)CCM_matrix[,i]$LibSize)
CCM_matrix_ColTarget <- sapply(1:ncol(CCM_matrix), function(i)CCM_matrix[,i]$`col:target`)
CCM_matrix_TargetCol <- sapply(1:ncol(CCM_matrix), function(i)CCM_matrix[,i]$`target:col`)
```

```{r fig.asp=0.3, fig.width=12}
par(mfrow=c(1,3))
matplot(CCM_matrix_LibSize, CCM_matrix_ColTarget, type="l", col=alpha("blue", 0.02), lwd=2, lty=1, xlab="Library size", ylab="Prediction Skill", main="Convergent cross mapping for origination")
lines(CCM_matrix_LibSize[,1], apply(CCM_matrix_ColTarget, 1, median), col="blue", lwd=2)
matplot(CCM_matrix_LibSize, CCM_matrix_TargetCol, type="l", col=alpha("red", 0.02), lwd=2, lty=1, add=T)
lines(CCM_matrix_LibSize[,1], apply(CCM_matrix_TargetCol, 1, median), col="red", lwd=2)
legend("topleft", legend=c("ori:ori", "ori:ori"), col=c("red", "blue"), lty=1, cex=0.8)

matplot(CCM_matrix_LibSize, CCM_matrix_ColTarget, type="l", col=alpha("blue", 0.02), lwd=2, lty=1, xlab="Library size", ylab="Prediction Skill", main="Convergent cross mapping for extinction")
lines(CCM_matrix_LibSize[,1], apply(CCM_matrix_ColTarget, 1, median), col="blue", lwd=2)
matplot(CCM_matrix_LibSize, CCM_matrix_TargetCol, type="l", col=alpha("red", 0.02), lwd=2, lty=1, add=T)
lines(CCM_matrix_LibSize[,1], apply(CCM_matrix_TargetCol, 1, median), col="red", lwd=2)
legend("topleft", legend=c("ext:ext", "ext:ext"), col=c("red", "blue"), lty=1, cex=0.8)

matplot(CCM_matrix_LibSize, CCM_matrix_ColTarget, type="l", col=alpha("blue", 0.02), lwd=2, lty=1, xlab="Library size", ylab="Prediction Skill", main="CCM for origination vs extinction")
lines(CCM_matrix_LibSize[,1], apply(CCM_matrix_ColTarget, 1, median), col="blue", lwd=2)
matplot(CCM_matrix_LibSize, CCM_matrix_TargetCol, type="l", col=alpha("red", 0.02), lwd=2, lty=1, add=T)
lines(CCM_matrix_LibSize[,1], apply(CCM_matrix_TargetCol, 1, median), col="red", lwd=2)
legend("topleft", legend=c("ext:ori", "ori:ext"), col=c("red", "blue"), lty=1, cex=0.8)
```

